\thetitle

### 2013-06-19 tfpie/tfpie.tex
up

\begin{abstract}
We propose a bridge between functional and object-oriented programming
in the first-year curriculum.  Traditionally, curricula that begin
with functional programming transition to a professional, usually
object-oriented, language in the second course.  This transition poses
numerous obstacles to students, and often results in confusing the
details of development environments, syntax, and libraries with the
fundamentals that the course focuses on.  Our proposal instead begins
the second course with a sequence of custom teaching languages which
minimize the transition from the first course, and allow students to
focus on the core ideas.  We then transition to Java half-way through
the course, at which point students have a strong command of the basic
ideas.  We have 3 years of experience with this course, and it has
produced notable success.
\end{abstract}

@@

Many universities and colleges aim to teach their students proficiency
in an industrial object-oriented programming language by the end of
the students' first year.  Past approaches to achieve this aim have
included teaching an industrial language, such as Java, starting on
the first day.  Others have taken a more indirect route by teaching
functional programming in the first semester, followed by a second
semester that starts in Java.  The latter approach is an improvement
over the first, but both suffer serious flaws.

As an example, Northeastern University teaches functional programming
in the first semester using \emph{How to Design
  Programs}~\cite{dvanhorn:Felleisen2001How}, followed by
object-oriented programming using \emph{How to Design
  Classes}~\cite{local:htdc} in the second semester.  This sequence
was designed to provide a smooth path for incoming students assumed
only to have a competence in high-school level algebra to reach
proficiency in Java by the end of their first year
\cite{dvanhorn:Felleisen2004Structure}.  It was a major improvement
over the previous Java-first curriculum in terms of student success,
attrition, and preparation for subsequent
courses~\cite{dvanhorn:Proulx2006Design}.  However, significant
problems remain; in particular, the second semester course violates
the designers' own principles (as recalled
in~\cite{dvanhorn:Bloch2000Scheme}):
%
\begin{enumerate}
\item \emph{introduce only those language constructs that are necessary to
  teach programming principles}, and
\item \emph{choose a language with as
  few language constructs as possible, and one in which they can be
  introduced one at a time}.
\end{enumerate}

The problem is that the first semester ends with an advanced
pedagogical functional language and the second semester starts with a
beginning pedagogical subset of Java.  Despite this slimmed-down
subset, this transition is too abrupt to meaningfully bridge the gap
between functional and object-oriented programming, because there are
several other significant transitions happening in concert such as the
transitions:
\begin{itemize}
\item from a highly regular and minimal syntax to a complicated
  irregular syntax,

@@

\item from a pedagogical programming environment to a professional
programming environment (Eclipse),

\item from a language with numeric values corresponding to
  mathematical objects to a language with numeric values corresponding
  to common machine representations,

@@

This abrupt transition has several negative consequences: the
principles of object-oriented programming are obscured and
de-emphasized, struggling with the programming environment is
frustrating and can cause potentially good students to leave the
program, it favors students with prior exposure to the particular
tools, it inhibits students from experimenting by relying upon their
past skills, and it creates the false impression that courses are
discrete units of instruction that can be discarded after successful
completion rather than being part of a continuous and cumulative
educational experience.

We contribute an alternative approach to the second semester that
overcomes these problems and provides a gradual transition from
functional programming to object-oriented programming.  Our approach
is able to start the second semester by introducing \emph{only} the
concept of programming with objects, while all other aspects of course
remain where they were left off in the previous semester.  This
approach allows the other concepts to be introduced at the point at
which they are relevant and motivated.  Despite this more gradual
approach, the course accomplishes the goal of teaching industrial
language proficiency by the end of the semester, covering a super-set
of the concepts and topics covered in the \emph{How to Design Classes}
course.

\paragraph{Outline} The remainder of this paper is organized as follows:
in section~\ref{sec:background} provides background on \emph{How to
  Design Programs} and the context and constraints involved in the
first-year at Northeastern.  Section~\ref{sec:shift} describes our
approach to the second semester, which starts with a small shift in
perspective to bridge the gap between functional programming and
object oriented programming.  Section~\ref{sec:industrial} describes
the path toward programming in an industrial object-oriented
programming language.  Section~\ref{sec:related-work} discusses the
relation to existing work and section~\ref{sec:conclusion} concludes.

>>>>

\begin{abstract}
We propose a bridge between functional and object-oriented programming
in the first-year curriculum.  Traditionally, curricula that begin
with functional programming transition to a professional, usually
object-oriented, language in the second course.  This transition poses
 obstacles for students, and often results in confusing the
details of development environments, syntax, and libraries with the
fundamentals of OO programming 
that the course focuses on.  Our proposal instead begins
the second course with a sequence of custom teaching languages which
minimize the transition from the first course, and allow students to
focus on core ideas.  We then transition to Java half-way through
the course, at which point students have a strong command of the basic
principles.  We have 3 years of experience with this course, with
 notable success.
\end{abstract}

@@

Many universities and colleges aim to teach their students proficiency
in an industrial object-oriented programming language by the end of
the students' first year.  The most common approache to achieve this
aim is to teach an industrial language in the first course, often
Java, starting on the first day.  Other curricula take a more indirect
route by teaching functional programming in the first semester,
followed by a second semester in Java.  The latter approach is an
improvement over the first, as pointed out by numerous
observers~\cite{dvanhorn:Felleisen2004Structure, local:spolsky}, but
both suffer serious flaws.

As an example, Northeastern University teaches introductory programming
in the first semester using \emph{How to Design
  Programs}~\cite{dvanhorn:Felleisen2001How}, followed by
object-oriented programming using \emph{How to Design
  Classes}~\cite{local:htdc} in the second semester.  This sequence
was designed to provide a smooth path for incoming students with
 a competence in high-school level algebra to reach
proficiency in Java by the end of their first year
\cite{dvanhorn:Felleisen2004Structure}.  It was a major improvement
over the previous Java-first curriculum in terms of student success,
attrition, and preparation for subsequent
courses~\cite{dvanhorn:Proulx2006Design}.  However, significant
problems remain; in particular, the second semester course violates
the designers' own principles (as recalled
in~\cite{dvanhorn:Bloch2000Scheme}):
%
\begin{enumerate}
\item \emph{introduce only those language constructs that are necessary to
  teach programming principles}, and
\item \emph{choose a language with as
  few language constructs as possible, and one in which they can be
  introduced one at a time}.
\end{enumerate}

The problem is that the first semester ends with an advanced
pedagogical functional language and the second semester starts with
Java, although it focuses on a side-effect free subset.  Despite this focused
subset, this transition is too abrupt to meaningfully bridge the gap
between functional and object-oriented programming, because
several other significant transitions happen in concert:
\begin{itemize}
\item from a highly regular and minimal syntax to a complicated
  irregular syntax,

@@

\item from a pedagogical programming environment (DrRacket) to a professional
programming environment (Eclipse),

\item from a language with numeric values corresponding to
  mathematical objects to a language with numeric values corresponding
  to common machine representations,\footnote{While this may seem like
    a minor point, details of numeric representation can crop up
    quickly in a Java-based course.}

@@

This abrupt transition has several negative consequences: the
principles of object-oriented programming are obscured and
de-emphasized, struggling with the programming environment is
frustrating and can cause potentially good students to leave the
program, it favors students with prior exposure to the particular
tools (a set that is demographically skewed), it inhibits students
from experimenting by allowing them to rely upon past skills, and it creates
the false impression that courses are discrete units of instruction
that can be discarded after successful completion rather than being
part of a continuous and cumulative educational experience.

We contribute an alternative approach to the second semester that
overcomes these problems and provides a gradual introduction 
to object-oriented programming.  Our approach
starts the second semester by introducing \emph{only} the
concept of programming with objects, while all other aspects of course
remain where they were left off in the previous semester.  This
 allows other concepts to be introduced at the point at
which they are relevant and motivated.  Despite this more gradual
approach, the course accomplishes the goal of reaching industrial
language competence by the end of the semester, covering a super-set
of the concepts and topics covered in the \emph{How to Design Classes}-based
course.

\paragraph{Outline} The remainder of this paper is organized as follows:
in section~\ref{sec:background} provides background on \emph{How to
  Design Programs} and the context and constraints involved in the
first year at Northeastern.  Section~\ref{sec:shift} describes our
approach to the second semester, which starts with a small shift in
perspective to bridge the gap between functional programming and
object oriented programming.  Section~\ref{sec:industrial} describes
the path from our pedagogical languages to an industrial object-oriented
programming language.  Section~\ref{sec:related-work} discusses the
relation to existing work and section~\ref{sec:conclusion} concludes.
